# Virtual Environments in R with `renv`
With `renv`:

- each project has its own package library
- package versions are recorded in a lockfile
- the same environment can be recreated later

## 3. Install `renv`

Run this once if `renv` is not installed yet.

```{r install-renv, eval=FALSE}
install.packages("renv")
```


Then load it:

```{r load-renv, eval=FALSE}
library(renv)
```


---

## 4. Create a new project environment

Move into your R project folder in RStudio or in your working directory, then initialize `renv`.

```{r renv-init, eval=FALSE}
renv::init()
```

What this does:

- creates a project-local library
- activates `renv` for the project
- creates supporting files, including `renv.lock`

After this step, your project is managed independently from your global R library.

## 5. Install packages inside the project

Now install the packages your project needs.

```{r install-project-packages, eval=FALSE}
install.packages("dplyr")
install.packages("ggplot2")
```


Because `renv` is active, these packages are tracked for this project.

You can check project status with:

```{r renv-status, eval=FALSE}
renv::status()
```

## 6. Save the environment

Once your project packages are installed, save the environment state.

```{r renv-snapshot, eval=FALSE}
renv::snapshot()
```

This updates the `renv.lock` file.

The lockfile records information such as:

- package names
- package versions
- repository sources

## 7. Restore the environment later

To recreate the same environment in the future or on another machine:

```{r renv-restore, eval=FALSE}
renv::restore()
```

This reads the `renv.lock` file and reinstalls the recorded package versions.

This is the main step that makes collaboration and reproducibility much easier.

## 8. Typical project workflow

A common `renv` workflow looks like this:

```{r workflow-example, eval=FALSE}
# one-time setup
install.packages("renv")
renv::init()

# work on the project
install.packages("readr")
install.packages("lubridate")

# save environment changes
renv::snapshot()

# later or on another machine
renv::restore()
```

## 9. Files to keep in version control

For a reproducible R project, you should usually commit:

- your `.R` or `.Rmd` files
- `renv.lock`
- the `renv/` infrastructure files that support the project

This allows another person to clone the project and run:

```{r restore-cloned-project, eval=FALSE}
renv::restore()
```

## 10. Common mistakes to avoid

Avoid these mistakes:

- using only the global R library
- forgetting to run `renv::snapshot()` after package changes
- sharing code without the lockfile
- deleting `renv.lock`
- assuming your collaborator has the same package versions

## 11. Small exercise 1

Create a new R project and do the following:

1. Initialize `renv`
2. Install `dplyr`
3. Install `ggplot2`
4. Run `renv::snapshot()`
5. Open the `renv.lock` file and identify the recorded packages

```{r exercise-restore, eval=FALSE}
renv::restore()
```